Scaffolder - Software for Manual Genome Scaffolding.

### 2011-09-12 src/article.tex
Proof read

  \paragraph*{Background:} The assembly of next-generation short-read
  sequencing data can result in a fragmented non-contiguous set of genomic
  sequences. Therefore a common step in a genome project is to join
  neighbouring sequence regions together and fill gaps . This scaffolding step
  is non-trivial and requires manually editing large blocks of nucleotide
  sequence. Joining these sequences together also hides the source of each
  region in the final genome sequence. Taken together, these considerations
  may make reproducing or editing an existing genome scaffold difficult.

@@

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent sequences using a relatively simple syntax. This
  syntax further allows unknown regions to be specified and additional sequence
  to be used fill known gaps in the scaffold. Defining the genome construction
  in a file makes the scaffolding process reproducible and easier to edit
  compared with large FASTA nucleotide sequences.

@@

The advent of high-throughput sequencing methods has led to a renewed interest
in algorithms to solve the problem of genome assembly \cite{pop2008,pop2009}.
The complexity of merging large numbers of overlapping reads can lead to
genome assembly software being unable to produce a complete sequence. Instead,
the algorithm may generate several large assembled regions of sequence
(`contigs') composed of the many individual readsThese contigs represent a
fragmented picture of the genome and therefore require additional work to join
together into a complete sequence. \pb

@@

The process of finishing a genome scaffold uses wet laboratory methods,
\emph{in silico} methods, or a combination of both. An example of a
computational method might use the paired-read data from the sequencing stage.
The occurrence of paired reads in separate contigs can be used to estimate
probabilistically the order and distance between these contigs. Alternatively,
laboratory methods may use PCR to amplify the unknown DNA in a gap region then
use Sanger sequencing to determine the sequence of this gap. Computational
methods are preferable as they are less costly in laboratory time and materials
compared to manual gap resolution, and use the available data that exists
following sequencing \cite{nagarajan2010}. Finally when the scaffold cannot be
completely resolved, \emph{in silico} software packages exist to suggest the
likely primers necessary for PCR amplifying the sequence in gap regions
\cite{gordon2001}. \pb

Examples of \emph{in silico} methods include comparing the assembled contigs
to a complete reference genome sequence to search for areas of sequence
similarity between the two. Any areas of corresponding sequence with the
reference genome can be used to infer contig placement and build the contigs
into a scaffold \cite{richter2007,zhao2008,assefa2009}. Genomic recombination
can however reduce the efficacy of this. \pb

@@

These described \emph{in silico} methods provide a wide array of approaches for
merging contigs into a larger, continuous scaffold sequence. The scaffolding
process may still require manually inserting additional sequences or further
joining contigs using PCR-derived sequence. Moving and editing large blocks of
nucleotide text by hand however introduces human error and precludes any
reproducibility of the steps in this process. \pb

@@

The choice of nucleotide sequences comprising the scaffold is specified using
the YAML syntax \cite{yaml}. YAML is a language format using whitespace and
indentation to produce a machine readable structure. As YAML is a standardised
data format, third-party developers have the option to generate a genome
scaffold using any programming language for which a YAML library exists. The
YAML website lists current parsers for languages including C/C++, Ruby, Python,
Java, Perl, C\#/.NET, PHP, and JavaScript. In addition to being widely
supported, YAML-formatted scaffold files can be validated for correct syntax
using third-party tools such as Kwalify \cite{kwalify}. \pb

@@

The input data for Scaffolder is nucleotide sequence in FASTA format file.
These nucleotide sequences can be of any length and for example may be
individual reads, assembled contigs or contigs which have been joined to into
larger scaffolds. The case in which Scaffolder may be most useful is using the
contigs and scaffolded contigs produced from the previously assembled set of
sequencing reads, combined with additional gap filling sequences produced by
PCR or \emph{in silico} methods as outlined in the Background. \pb

@@

The Scaffolder software facilitates reproducibly joining nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in YAML specify how these sequences should be joinedThe
Scaffolder software is used to generate the scaffold sequence from these
instructions. In addition to specifying which contigs are required, the
scaffold file allows the contigs to be edited into smaller sub-sequences or
reverse complemented if necessary. Each scaffold file represents one scaffolded
nucleotide sequence and separate scaffolds should be defined in separate files.
\pb

@@

Scaffolder uses a minimal and compact syntax to describe how a genome scaffold
sequence should be generated. This syntax is simple to write and edit whilst
being succinct and readable.

AGP is a similar format for describing scaffolds. This format can be used to
describe contig order and N-filled gap regions in a scaffold. The advantages of
the AGP format is that each contig entry is defined on a single line which
allows searching the scaffold using UNIX line-based tools. The Scaffolder
format in comparison in contrast is written in the standardised YAML format and
therefore accessible to the many languages which provide parsers to this
format. The Scaffolder format is provided with a tool explicitly to produce the
fasta sequence of scaffold specified by in the file. The scaffold format
further provides functions for trimming and replacing regions of sequence using
inserts. \pb

@@

Scaffolder is software aimed at both bioinformaticians and biologists familiar
with the command line who wish to build a genome scaffold from a set of
contigs. The Scaffolder file format maintains the genome scaffold as a concise
and readable text representation that allows third parties to see how the
genome sequence was scaffolded. This file format also allows a broad overview
of which sequences were included and how they are ordered in the genome
scaffold, something not possible to deduce from a megabase-length string of
nucleotide characters. Scaffolder furthers increases the ease of
reproducibility in genome projects by allowing the scaffold super-sequence to
be reliably reproduced from the same scaffold file. The YAML syntax for writing
the scaffold file is also standardised and simple to manipulate
programmatically. This thereby means the scaffolding process follows the Unix
tenet of ``If your data structures are good enough, the algorithm to manipulate
them should be trivial.''

>>>>

  \paragraph*{Background:} The assembly of next-generation short-read
  sequencing data can result in a fragmented non-contiguous set of genomic
  sequences. Therefore a common step in a genome project is to join
  neighbouring sequence regions together and fill gaps . This scaffolding step
  is non-trivial and requires manually editing large blocks of nucleotide
  sequence. Joining these sequences together also hides the source of each
  region in the final genome sequence. Taken together these considerations
  may make reproducing or editing an existing genome scaffold difficult.

@@

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent sequences using a relatively simple syntax. This
  syntax further allows unknown regions to be specified and additional sequence
  to be used to fill known gaps in the scaffold. Defining the genome construction
  in a file makes the scaffolding process reproducible and easier to edit
  compared with large FASTA nucleotide sequences.

@@

The advent of high-throughput sequencing methods has led to a renewed interest
in algorithms to solve the problem of genome assembly \cite{pop2008,pop2009}.
The complexity of merging large numbers of overlapping reads can lead to
genome assembly software being unable to produce a complete sequence. Instead,
the algorithm may generate several large assembled regions of sequence
(`contigs') composed of the many individual reads. These contigs represent a
fragmented picture of the genome and therefore require additional work to join
together into a complete sequence. \pb

@@

The process of finishing a genome scaffold uses wet laboratory methods,
\emph{in silico} methods, or a combination of both. An example of a
computational method might use the paired-read data from the sequencing stage.
The occurrence of paired reads in separate contigs can be used to estimate
probabilistically the order and distance between these contigs. Alternatively,
laboratory methods may use PCR to amplify the unknown DNA in a gap region then
use Sanger sequencing to determine the sequence of this gap. Computational
methods using the available sequencing data are preferable as they are less
costly in laboratory time and materials compared to manual gap resolution
\cite{nagarajan2010}. Finally when the scaffold cannot be completely resolved,
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions \cite{gordon2001}. \pb

Examples of \emph{in silico} methods include comparing the assembled contigs
to a complete reference genome sequence to search for areas of sequence
similarity between the two. Any areas of corresponding sequence in the
reference genome can be used to infer contig placement and build the contigs
into a scaffold \cite{richter2007,zhao2008,assefa2009}. Genomic recombination
can however reduce the efficacy of this. \pb

@@

These described \emph{in silico} methods provide a wide array of approaches for
merging contigs into a larger, continuous scaffold sequence. The scaffolding
process may still require manually inserting additional sequences or further
joining contigs using PCR-derived sequence. Moving and editing large blocks of
nucleotide text by hand however introduces human error and precludes any
reproducibility. \pb

@@

The choice of nucleotide sequences comprising the scaffold is specified using
the YAML syntax \cite{yaml}. YAML is a data format using whitespace and
indentation to produce a machine readable structure. As YAML is a standardised
data format, third-party developers have the option to generate a genome
scaffold using any programming language for which a YAML library exists. The
YAML website lists current parsers for languages including C/C++, Ruby, Python,
Java, Perl, C\#/.NET, PHP, and JavaScript. In addition to being widely
supported, YAML-formatted scaffold files can be validated for correct syntax
using third-party tools such as Kwalify \cite{kwalify}. \pb

@@

The input data for Scaffolder is nucleotide sequence in FASTA format file.
These nucleotide sequences can be of any length and for example may be
individual reads, assembled contigs or contigs which have been joined to into
larger scaffolds. The case in which Scaffolder may be most useful is using the
contigs and scaffolded contigs, combined with additional gap filling sequences
produced by PCR or \emph{in silico} methods as outlined in the Background. \pb

@@

The Scaffolder software facilitates reproducibly joining nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in YAML specify how these sequences should be joined.
The Scaffolder software is used to generate the scaffold sequence from these
instructions. In addition to specifying which contigs are required, the
scaffold file allows the contigs to be edited into smaller sub-sequences or
reverse complemented if necessary. Each scaffold file represents one scaffolded
nucleotide sequence and separate scaffolds should be defined in separate files.
\pb

@@

Scaffolder uses a minimal and compact syntax to describe how a genome scaffold
sequence should be generated. This syntax is simple to write and edit whilst
being succinct and readable. \pb

AGP is a similar format for describing scaffolds. This format can be used to
describe contig order and N-filled gap regions in a scaffold. The advantages of
the AGP format is that each contig entry is defined on a single line which
allows searching the scaffold using UNIX line-based tools. The Scaffolder
format in comparison in contrast is written in the standardised YAML format and
therefore accessible to the many languages which provide parsers to this
format. The Scaffolder format is provided with a tool explicitly to produce the
FASTA sequence of scaffold specified by in the file. The Scaffolder format
further provides functions for trimming and replacing regions of sequence using
inserts. \pb

@@

Scaffolder is software aimed at both bioinformaticians and biologists familiar
with the command line who wish to build a genome scaffold from a set of
contigs. The Scaffolder file format maintains the genome scaffold as a concise
and readable text representation that allows third-parties to see how the
genome sequence was scaffolded. This file format also allows a broad overview
of which sequences were included and how they are ordered in the genome
scaffold, something not possible to deduce from a megabase-length string of
nucleotide characters. Scaffolder furthers increases the ease of
reproducibility in genome projects by allowing the scaffold super-sequence to
be reliably reproduced from the same scaffold file. The YAML syntax for writing
the scaffold file is also standardised and simple to manipulate
programmatically. This thereby means the scaffolding process follows the Unix
tenet of ``If your data structures are good enough, the algorithm to manipulate
them should be trivial.''
